\name: A flexible and data-driven $K$-process model for nucleosynthesis

### 2023-06-27 Paper/main.tex
incorporating comments from readers

% Abundance and stellar macros
\newcommand{\mgfe}[0]{[{\rm Mg/Fe}]}
\newcommand{\afe}[0]{[\alpha/{\rm Fe}]}
\newcommand{\femg}{[{\rm Fe}/{\rm Mg}]} 
\newcommand{\xmg}{[{\rm X}/{\rm Mg}]} 
\newcommand{\mgh}{[{\rm Mg}/{\rm H}]}
\newcommand{\feh}[0]{[{\rm Fe/H}]}
\newcommand{\xfe}{[{\rm X}/{\rm Fe}]}
\newcommand{\xh}{[{\rm X}/{\rm H}]}
\newcommand{\logg}{\log(g)}
\newcommand{\teff}{T_{\rm eff}}
\newcommand{\kpc}{\rm \; kpc}
\newcommand{\kel}{\rm \; K}
\newcommand{\msun}{M_{\odot}}
\newcommand{\rgc}{R_{\rm GC}}
\newcommand{\msunvice}{M_{\odot}/M_{\odot \rm formed}}
\newcommand{\MZAMS}{M_{\rm ZAMS}}
\newcommand{\qcc}{q_{{\rm CC,}j}^{Z}}
\newcommand{\qccFe}{q_{{\rm CC, Fe}}^{Z}}
\newcommand{\dqccFe}{dq_{{\rm CC, Fe}}^{Z}/dZ}
\newcommand{\qIa}{q_{{\rm Ia,}j}^{Z}}
\newcommand{\qIaFe}{q_{{\rm Ia,Fe}}^{Z}}
\newcommand{\Acc}{A^{\rm CC}_{i}}
\newcommand{\AIa}{A^{\rm Ia}_{i}}
\newcommand{\fcc}{f_{{\rm CC}, ij}}

\newcommand{\ejg}[1]{\textcolor{red}{EJG says: #1}}
\newcommand{\hogg}[1]{\textcolor{blue}{Hogg says: #1}}

@@

\begin{abstract}\noindent % trust
Stellar surface abundances look like they are produced by two dominant processes, one prompt and one delayed.
We analyze the abundances of 14 elements in [X/Mg] vs. [Mg/H] space for 48,659 stars from APOGEE DR17 with a flexible, data-driven $K$-process model---dubbed \name.
In the fiducial model, where $K=2$, each elemental abundance of each star is described as the vector sum of a prompt (CCSN-like) and delayed (SNIa-like) process.
While prior work derives process vector components from median observed high-Ia and low-Ia abundance trends, we relax these assumptions and simultaneously fit all model parameters to the data, with only minimal constraints to keep the processes interpretable.
We find that our flexible method is able to recover the observed abundance bimodality and that this bimodality is visible to higher [Mg/H] in process amplitude space than in abundance space.
We compare fit parameters and statistics from this new method to those from prior work, finding that the two methods produce similar results, but that \name{} better predicts stellar abundances, especially for C+N and Mn.
We test the validity of the standard two-process model assumptions, such as that Fe is generated by a roughly half-half combination of prompt and delayed sources.
We find that the assumptions of prior works are generally confirmed, but that the inferred fractional contributions from each process are strongly dependent upon Fe assumptions.
We also explore third and fourth processes, calibrated to Ce and Mn, respectively. The additional processes improve the model's ability to predict the stellar abundances, demonstrating the model's ability to extend beyond $K=2$. \name{} enables future research to study the disk's two-dimensionallity, investigate age-abundance label correlations, fit non-bimodal abundance distributions with a multi-component nucleosynthetic model, and much more. \ejg{Add something about code availability.} 
\end{abstract}

@@

After hydrogen, helium, lithium, and beryllium, all other naturally occurring elements are made in stars, supernovae, and the collisions of stars.
Stellar surface abundances---the abundances measured by taking a spectrum of a stellar photosphere---are thought to deliver a relatively unprocessed record of the element abundances in the gas from which the star formed \citep[though see e.g.,][]{pinsonneault2001, oh2018, souto2019, vincenzo2021b}.
These birth abundances were set by a combination of nucleosynthetic processes involved in making heavy atomic nuclei, and astrophysical processes involved in delivering atoms from stellar interiors to star-formation sites \citep[e.g.,][]{johnsonja2020}.
Thus nuclear physics and a wide swath of astrophysics are critically intertwined in our understanding of stellar surface abundances, motivating theoretical, experimental, and observational work.

At the present day, stellar surface abundances are not very well explained by \textsl{ab initio}, physics-driven models.
Yields vary from set to set, as they are dependent on progenitor properties and explosion assumptions \citep[e.g.,][]{rybizki2017, griffith2021b}. 
The wide parameter space of progenitor and supernovae models coupled with uncertainties in reaction rates and explosion physics hinder the creation of an accurate nucleosynthetic model from theory alone.
In the long run, it is incumbent upon us to understand these issues and correct the assumptions or calculations underlying our nucleosynthetic and astrophysical models.
In the short run, however, we gather data---tens of millions of abundance measurements on millions of stars in different astronomical surveys such as RAVE, SEGUE, LAMOST, Gaia-ESO, APOGEE/MWM, GALAH, and H3 \citep{steinmetz2006, yanny2009, gilmore2012, desilva2015, luo2015, majewski2017, conroy2019}.
This raises the question: \emph{Can we take a data-driven approach to nucleosynthesis?}

In this \documentname{}, we build a purely data-driven model for the surface element abundances observed in stars.
We treat each star as being a linear combination of nucleosynthetic processes, beginning with one that is primarily responsible for the $\alpha$-element Mg \citep[prompt enrichment, such as core collapse supernovae or CCSN, e.g.,][]{andrews2017}, and one that is primarily \emph{not} responsible for Mg (delayed enrichment, such as Type-Ia supernovae or SNIa).
Beyond these up-front assumption, we try to be agnostic about how the elements are produced.

We build upon the work of \citet[][hereafter G22]{griffith2019, griffith2022} and \citet[][hereafter W19, W22]{weinberg2019, weinberg2022}, who used the bimodality in [Mg/Fe] vs. [Fe/H]\footnote{Where $[{\rm X}/{\rm Y}] = \log({\rm X}/{\rm Y}) - \log({\rm X}/{\rm Y})_{\odot}$ and $({\rm X}/{\rm Y})_{\odot}$ is the solar abundance ratio.} \citep[e.g.,][]{fuhrmann1998, bensby2003, adibekyan2012} to separate stars into populations with high and low SNIa enrichment. Using the median [X/Mg] vs. [Mg/H] abundance trends, these works explain data from the \textsl{GALAH}\footnote{GALAH = GALactic Archaeology with HERMES.} and SDSS-IV \textsl{APOGEE}\footnote{APOGEE = Apache Point Observatory Galactic Evolution Experiment, part of the Sloan Digital Sky Survey} surveys, respectively, with a two-process model. Because the median abundance trends in [X/Mg] vs. [Mg/H] space are largely insensitive to aspects of chemical evolution, such as outflows and variations in star formation history \citep{weinberg2019}, the population abundance trends are set by the nucleosynthetic processes and can be used to empirically constrain Galactic enrichment.

These works, as well as \citet{ting2022}, find that the Milky Way stellar abundances are well fit two components, down to residuals of 0.01 to 0.02 dex. Simultaneously, \citet{frankel2018} and \citet{ness2022} have found that disk abundances are also well described by a two component model of birth radius and age. Correlations between two-process model parameters and stellar ages and kinematics (W22) suggest that these two 2-dimensional models are somehow interconnected. \ejg{Add bit on why two processes dominate---two main channels v.s. mixing in the disk.}

Beyond standard CCSN and SNIa enrichment many elements have contributions from additional nucleosynthetic processes, such as the rapid ($r$) and slow ($s$) neutron capture processes \citep[e.g.,][]{arlandini1999, bisterzo2014} in asymptotic giant branch (AGB) stars \citep[e.g.,][]{simmerer2004, karakas2016} and merging neutron stars \citep[e.g.,][]{kilpatrick2017}, or atypical supernovae explosion \citep[e.g.,][]{nomoto2013}. After predicting stellar abundances from $\feh$ and $\mgfe$, \citet{ting2022} identify correlated abundance residuals that are unexplained by observational uncertainties, indicative of additional nucleosynthetic processes that standard disk CCSN and SNIa enrichment cannot explain. Results from G22 and W22 support this conclusion, and both works attempt to add additional processes to their models to account for non-CCSN and non-SNIa enrichment, though in a very restrictive manner. Other sources of abundance scatter, such as stochastic sampling of the IMF, IMF variations, and bursty star formation history could also cause deviations away from a two-process model \citep{belokurov2018, griffith2023}.

To date, survey abundances have not been fully exploited to create a data-driven model of nucleosynthesis. While works such as \citet{ting2012}, \citet{casey2019}, and \citet{ratcliffe2022} effectively use clustering algorithms to identify elements with like sources and reduce abundance dimensionality, the results are difficult to translate into a model of nucleosynthesis. Clustering components can be linked to nucleosynthesis sources and enrichment history, but have not yet been be used to describe the enrichment of a single star.

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data.
In the $K$ Process Model (\name{}), we find the intersection between reliable facts about nucleosynthesis and good abundance measurements to build an edifice of Galactic enrichment.
The model is hierarchical, in that it learns some parameters (process vectors) that are shared across all stars, but different for each element, and some parameters (process amplitudes) that are shared across all elements, but different for each star.
The parameters output by our model can thus be used as de-noised abundance labels; these will sharpen relationships between abundances and stellar parameters (including birth location and time).

@@

As in W22 and G22, we propose that all stellar abundances can be generated by a combination of $K$ nucleosynthetic processes.
In this picture, each element has $K$ metallicity-dependent process vector components that are shared across the full stellar sample, while each star individually has $K$ process amplitudes, which apply across all elements, such that
\begin{equation}\label{eq:mij_k}
    m_{ij} = \log_{10} \, \sum^K_{k=1} A_i^k \, q_{k,j}^Z ~,
\end{equation}
where stars are labeled by $i$ and elements are labeled by $j$.
Each star $i$ has $K$ process amplitudes ($A^k_i$)
and each element $j$ has $K$ process vector components ($q_{k,j}^{Z}$).
The $m_{ij}$ value is the model expected logarithmic abundance of element $j$ in star $i$. In this model, we put forth the following set of assumptions:

\paragraph{1. $K$ processes}
All elements on the periodic table are produced by a combination of nucleosynthetic processes such as CCSN, SNIa, AGB stars, and merging neutron stars \citep{johnsonja2020}. The majority of $\alpha$, light odd-Z, and Fe-peak elements (the elements observed by APOGEE) are dominantly produced by $K=2$ sources, with one being a prompt process or mix of prompt processes, and one being a delayed process or a mix of delayed processes. This is substantiated by theoretical yields \citep[e.g.,][]{anderson2019, rybizki2017} and past successful data-driven models (e.g., G22, W22, \citealp{ting2022}). In this paper we therefore assume that $K \geq 2$, though \name{} could in principle be implemented with $K=1$.

@@

In \name{}, this assumption is enforced by fixing the Mg process vector components such that
\begin{equation}\label{eq:qcc_mg_solar}
    q_{{\rm CC, Mg}}^{\,Z_{\odot}} = 1, \quad q_{k>1,{\rm Mg}}^{\,Z_{\odot}} = 0
\end{equation}
at solar metallicity and 
\begin{equation}\label{eq:qcc_mg_z}
    q_{{\rm CC, Mg}}^{\,Z} = q_{{\rm CC, Mg}}^{\,Z_{\odot}}, \quad q_{k>1, {\rm Mg}}^{\,Z} = q_{k>1, {\rm Mg}}^{\,Z_{\odot}}
\end{equation}
at all other metallicities.

\paragraph{5. Fe production}
Fe is produced through a combination of a prompt and delayed process. Because the delayed process is likely dominated by SNIa \citep[e.g.,][]{andrews2017}, we label the delayed process with ``Ia''\footnote{While other enrichment channels with similar timescales may be included in the respective processes, the ``CC'' and ``Ia'' naming convention conforms to the choices in W22 and G22, and avoids the possible confusion of process numbers (1 and 2) with supernovae type (II and Ia).}.
While the prompt process constraint (Mg) is grounded in nucleosynthesis theory, there is no equivalent nucleosynthesis fact to constrain the delayed process.
To break model degeneracies, we also fix the Fe process vector components such that

\begin{equation}\label{eq:qcc_fe_solar}
    q_{{\rm CC, Fe}}^{\,Z_{\odot}} = 0.4, \quad q_{{\rm Ia, Fe}}^{\,Z_{\odot}} = 1 - q_{{\rm CC, Fe}}^{\,Z_{\odot}} \quad q_{k>2,{\rm Fe}}^{\,Z_{\odot}} = 0
\end{equation}
at solar metallicity and 
\begin{equation}\label{eq:qcc_fe_z}
    q_{{\rm CC, Fe}}^{\,Z} = q_{{\rm CC, Fe}}^{\,Z_{\odot}},  \quad 
    q_{{\rm Ia, Fe}}^{\,Z} = q_{{\rm Ia, Fe}}^{\,Z_{\odot}} \quad 
    q_{k>2, {\rm Fe}}^{\,Z} = q_{k>2, {\rm Fe}}^{\,Z_{\odot}}
\end{equation}
at all other metallicities.
This assumption places a star with purely prompt enrichment on the low-metallicity $\mgfe$ plateau near 0.4 dex, in agreement with APOGEE observations but in contention with recent results from \citet{conroy2022} which places the $\mgfe$ plateau near 0.6 dex.
We explore the impact of different $\qccFe$ assumptions in Section~\ref{subsec:qccFe}.

@@

\paragraph{7. APOGEE abundances and uncertainties}
We assume that the APOGEE abundances and uncertainties can be used for this project.
This is not the same as assuming that they are correct, but rather that it is possible and useful to build an interpretable model to explain them. We describe the potential data systematics in Section~\ref{sec:data}.
The APOGEE errors may be slightly underestimated for some element measurements, so we add a softening parameter $Q$ to inflate the uncertainties on abundance-outlier stars (see Equation~\ref{eq:inflate_ivar}).

\paragraph{8. Robust likelihood function}
The observed value of [X/H] can be described as the $K$ process expectation plus observational noise and/or other sources of intrinsic abundance scatter that are not included in this model:
\begin{equation} \label{eq:xh}
    \xh_{ij} = m_{ij} + \text{noise} ~,
\end{equation}
where $m_{ij}$ is the expectation value given in Equation \ref{eq:mij_k} for star $i$ and element $j$.
The expression in Equation \ref{eq:xh} can be thought of as the key assumption underlying our likelihood function.
In detail the (negative two times the) log likelihood function is given by a chi-squared ($\chi^2$) objective
\begin{equation}
    \chi^2 = \sum_{ij} \, \frac{1}{\sigma_{ij}^2} \, (\xh_{ij} - m_{ij})^2 + Q ~,
\end{equation}
where $1/\sigma_ij^2$ is the (robust; see below) inverse variance on measurement $ij$,
and $Q$ is a light regularization that depends on the parameters.

Because we don't want to be too drawn or influenced by outlier points, we don't use the observed errors $\sigma_{{\rm obs},ij}$ in the likelihood, but instead we soften them in the spirit of iteratively reweighted least squares (\citealt{irls}):
\begin{equation}\label{eq:inflate_ivar}
    \frac{1}{\sigma_{ij}^2} = \frac{Q^2/\sigma_{{\rm obs},ij}^2}{Q^2 + (\xh_{ij} - m_{ij})^2 /\sigma_{{\rm obs},ij}^2 },    
\end{equation}
where $Q$ is a softening parameter, which is set to $Q=5$.

\paragraph{9. Implementation and optimization} 
With the above assumptions in place, the likelihood function can be optimized to a set of stellar abundances.
The model is initialized at the Mg and Fe process vector components from Equations~\ref{eq:qcc_mg_solar} --~\ref{eq:qcc_fe_z}. It subsequently optimizes the process amplitudes (dubbed the $A$-step) using only Mg and Fe at fixed process vector components, and then optimizes the process vector components (dubbed the $q$-step) for all elements at fixed process amplitudes.
The $A$-step and $q$-step are alternated, repeating 48 rounds of optimization and updating the best-fit parameters when the objective function improves. We find few differences in the best-fit parameters when we decrease the number of iterations to 32, indicating that the model quickly finds a good solution.  
In detail the optimizations are performed with a nonlinear $\chi^2$ minimization algorithm (Gauss--Newton nonlinear least-squares) from \texttt{jaxopt}\footnote{\url{https://jaxopt.github.io/}}.

\bigskip
\name{} mirrors the two-proccess model from prior work (G22, W22) but, unless otherwise noted, the assumptions weaker, there is a likelihood function in play, and the implementation is more general.
In particular, we don't assume anything about the relationships between the process vector components and the morphologies of observed element-abundance ratio diagrams.

>>>>

% Abundance and stellar macros
\newcommand{\mgfe}[0]{[{\rm Mg/Fe}]}
\newcommand{\afe}[0]{[\alpha/{\rm Fe}]}
\newcommand{\femg}{[{\rm Fe}/{\rm Mg}]} 
\newcommand{\xmg}{[{\rm X}/{\rm Mg}]} 
\newcommand{\mgh}{[{\rm Mg}/{\rm H}]}
\newcommand{\feh}[0]{[{\rm Fe/H}]}
\newcommand{\xfe}{[{\rm X}/{\rm Fe}]}
\newcommand{\xh}{[{\rm X}/{\rm H}]}
\newcommand{\logg}{\log(g)}
\newcommand{\teff}{T_{\rm eff}}
\newcommand{\kpc}{\rm \; kpc}
\newcommand{\kel}{\rm \; K}
\newcommand{\msun}{M_{\odot}}
\newcommand{\rgc}{R_{\rm GC}}
\newcommand{\msunvice}{M_{\odot}/M_{\odot \rm formed}}
\newcommand{\MZAMS}{M_{\rm ZAMS}}
\newcommand{\qcc}{q_{{\rm CC,}j}^{Z}}
\newcommand{\qccFe}{q_{{\rm CC, Fe}}^{Z}}
\newcommand{\dqccFe}{dq_{{\rm CC, Fe}}^{Z}/dZ}
\newcommand{\qIa}{q_{{\rm Ia,}j}^{Z}}
\newcommand{\qIaFe}{q_{{\rm Ia,Fe}}^{Z}}
\newcommand{\Acc}{A^{\rm CC}_{i}}
\newcommand{\AIa}{A^{\rm Ia}_{i}}
\newcommand{\fcc}{f_{ij}^{\rm CC}}

\newcommand{\qccX}[1]{q_{{\rm CC, #1}}^{Z}}
\newcommand{\qIaX}[1]{q_{{\rm Ia, #1}}^{Z}}
\newcommand{\fccX}[1]{f_{i, {\rm #1}}^{\rm CC}}

\newcommand{\ejg}[1]{\textcolor{red}{EJG says: #1}}
\newcommand{\add}[1]{\textcolor{teal}{#1}}

\newcommand{\hogg}[1]{\textcolor{blue}{Hogg says: #1}}

@@

\author{Julianne Dalcanton}
\author{Sten Hasselquist}
\author{Bridget Ratcilffe}
\author{Melissa Ness}
\author{David Weinberg}

\begin{abstract}\noindent % trust
\add{The origin of the surface abundance patterns of Milky Way stellar populations can be explained by two dominant processes,} one prompt and one delayed.
\add{While two components can well describe the stellar abundance trends, elements are known to be produced by more than two sources.}
We analyze the abundances of 14 elements relative to Mg for 48,659 stars from APOGEE DR17 with a flexible, data-driven $K$-process model---dubbed \name.
In the fiducial model, where $K=2$, each elemental abundance of each star is described as the vector sum of a prompt and delayed process.
%While prior work derives the relative importance of \add{the prompt and delayed processes in the production of various elements} from median observed abundance trends, we relax these assumptions and 
%We simultaneously fit all model parameters to the data, with only minimal constraints to keep the processes interpretable.
We find that our flexible method is able to recover the observed abundance bimodality and that this bimodality is visible to higher [Mg/H] in process amplitude space than in abundance space.
We compare fit parameters and statistics from this new method to those from \citet{weinberg2022}, finding that the two methods produce similar results, but that \name{} better predicts stellar abundances, especially for C+N and Mn.
%We test the validity of the standard two-process model assumptions, such as that in the sun Fe is generated by a roughly half-half combination of prompt and delayed sources.
%We find that the assumptions of prior works are generally confirmed, but that the inferred fractional contributions from each process are strongly dependent upon Fe assumptions.
\add{We explore the impact of our assumptions about the origin of Fe, needed to break the rotational symmetry of the model, constraining the parameter space that is able to reproduce the observed abundance distribution and finding that the inferred fractional contribution from each process are dependent upon our choice.}
Finally, we add a third and fourth process, calibrated to Ce and Mn, respectively. The additional processes \add{give the model more freedom} and improve the model's ability to predict the stellar abundances, \add{as expected}, but demonstrates the model's ability to extend beyond $K=2$. 
\name{} enables future research to study the disk's abundance dimensionallity, investigate age-abundance label correlations, fit non-bimodal abundance distributions with a multi-component nucleosynthetic model, and much more. 
\ejg{Hogg: try to shorten the abstract. I made some changes but still need to cut like 25 more words.} 
\end{abstract}

@@

After hydrogen, helium, lithium, and beryllium, all other naturally occurring elements are made in stars, supernovae, and the collisions of stars.
Stellar surface abundances---the abundances measured by taking a spectrum of a stellar photosphere---are thought to deliver a relatively unprocessed record of the element abundances in the gas from which the star formed \citep[though see, e.g.,][]{pinsonneault2001, oh2018, souto2019, vincenzo2021b}.
These birth abundances were set by a combination of nucleosynthetic processes involved in making heavy atomic nuclei, and astrophysical processes involved in delivering atoms from stellar interiors to star-formation sites \citep[e.g.,][]{johnsonja2020}.
Thus nuclear physics and a wide swath of astrophysics are critically intertwined in our understanding of stellar surface abundances, motivating theoretical, experimental, and observational work.

At the present day, stellar surface abundances are not very well explained by purely \textsl{ab initio}, physics-driven models.
Theoretical yields vary from set to set, as they are dependent on progenitor properties and explosion assumptions \citep[e.g.,][]{rybizki2017, blancato2019, buck2021, griffith2021b}. 
The wide parameter space of progenitor and supernova models coupled with uncertainties in reaction rates and explosion physics hinder the creation of an accurate nucleosynthetic model from theory alone.
In the long run, it is incumbent upon us to understand these issues and correct the assumptions or calculations underlying our nucleosynthetic and astrophysical models.
In the short run, however, we gather data---tens of millions of abundance measurements on millions of stars in different astronomical surveys such as RAVE, SEGUE, LAMOST, Gaia-ESO, APOGEE/MWM, GALAH, and H3 \citep{steinmetz2006, yanny2009, gilmore2012, desilva2015, luo2015, majewski2017, conroy2019}.
This raises the question: \emph{Can we take a data-driven approach to nucleosynthesis?}

In this \documentname{}, we build a purely data-driven model for the surface element abundances observed in stars.
We treat each star as being a linear combination of nucleosynthetic processes, beginning with one that is primarily responsible for the $\alpha$-element Mg \citep[prompt enrichment, such as core-collapse supernovae or CCSN, e.g.,][]{andrews2017}, and one that is primarily \emph{not} responsible for Mg (delayed enrichment, such as Type-Ia supernovae or SNIa).
Beyond these up-front assumptions, we try to be agnostic about how the elements are produced.

We build upon the work of \citet[][hereafter G22]{griffith2019, griffith2022} and \citet[][hereafter W19, W22]{weinberg2019, weinberg2022}, who used the bimodality in [Mg/Fe] vs. [Fe/H]\footnote{Where $[{\rm X}/{\rm Y}] = \log({\rm X}/{\rm Y}) - \log({\rm X}/{\rm Y})_{\odot}$ and $({\rm X}/{\rm Y})_{\odot}$ is the solar abundance ratio.} \citep[e.g.,][]{fuhrmann1998, bensby2003, adibekyan2012} to separate stars into populations with high and low SNIa enrichment. Using the median [X/Mg] vs. [Mg/H] abundance trends, these works explain data from the GALAH\footnote{GALAH = GALactic Archaeology with HERMES.} and SDSS-IV APOGEE\footnote{APOGEE = Apache Point Observatory Galactic Evolution Experiment, part of the Sloan Digital Sky Survey} surveys, respectively, with a two-process model. Because the median abundance trends in [X/Mg] vs. [Mg/H] space are largely insensitive to aspects of chemical evolution, such as outflows and variations in star formation history \citep{weinberg2019}, the population abundance trends are set by the nucleosynthetic processes and can be used to empirically constrain Galactic enrichment.

These works, as well as \citet{ting2022} and \citet{ratcliffe2023}, find that the Milky Way stellar abundances are well fit by two components, \add{grounded in [Fe/H] and [Mg/Fe],} down to residuals of 0.01 to 0.02 dex. Simultaneously, \citet{frankel2018} and \citet{ness2022} have found that disk abundances are also well described by a two-component model of birth radius and age. Correlations between two-process model parameters and stellar ages and kinematics (W22) \add{as well as the success of a two-component model of [Fe/H] and age in predicting APOGEE abundances \citep{ness2019}} suggest that these two 2-dimensional models are somehow interconnected.

Beyond standard CCSN and SNIa enrichment, many elements have contributions from additional nucleosynthetic processes, such as the rapid ($r$) and slow ($s$) neutron capture processes \citep[e.g.,][]{arlandini1999, bisterzo2014} in asymptotic giant branch (AGB) stars \citep[e.g.,][]{simmerer2004, karakas2016} and merging neutron stars \citep[e.g.,][]{kilpatrick2017}, or atypical supernova explosions \citep[e.g.,][]{nomoto2013}. After predicting stellar abundances from $\feh$ and $\mgfe$, \citet{ting2022} identify correlated abundance residuals that are unexplained by observational uncertainties, indicative of additional nucleosynthetic processes that standard disk CCSN and SNIa enrichment cannot explain. Results from G22 and W22 support this conclusion, and both works attempt to add additional processes to their models to account for non-CCSN and non-SNIa enrichment, though in a very restrictive manner. Other sources of abundance scatter, such as stochastic sampling of the Initial Mass Function (IMF), IMF variations, and bursty star formation history could also cause deviations away from a two-process model \citep{belokurov2018, griffith2023}.

To date, survey abundances have not been fully exploited to create a data-driven model of nucleosynthesis. While works such as \citet{ting2012}, \citet{casey2019}, and \citet{ratcliffe2020} effectively use clustering algorithms to identify elements with like sources and reduce abundance dimensionality, the results are difficult to translate into a model of nucleosynthesis. Clustering components can be linked to nucleosynthesis sources and enrichment history, but have not yet been used to describe the enrichment of a single star.

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data.
In the $K$ Process Model (\name{}), we find the intersection between reliable facts about nucleosynthesis and good abundance measurements to build an edifice of Galactic enrichment.
The model is hierarchical, in that it learns some parameters (process vectors) that are shared across all stars, but different for each element, and some parameters (process amplitudes) that are shared across all elements, but different for each star.
The parameters output by our model can thus be used as de-noised abundance labels; these will sharpen relationships between abundances and stellar parameters (including birth location and time). \add{Our main contribution is to construct a data-driven model for nucleosynthesis that has good statistical properties, only enforcing constraints to break rotational degeneracies. All other \name{} parameters are set by the data with no fixed normalization.}
% Had said some of this in the sentences before...
%The process parameters are derived Uses a hierarchical statistical model so it combines abundances into K process amplitudes for each star and denoised predicted abundances of each star, and metallicity-dependent effective yields for the processes.

@@

As in W22 and G22, we propose that all stellar abundances can be generated by a combination of $K$ nucleosynthetic processes.
In this picture, each element has $K$ metallicity-dependent process vector components that are shared across the full stellar sample, while each star individually has $K$ process amplitudes, \add{which apply across all elements, such that the expected logarithmic abundance of element $j$ in star $i$ ($m_{ij}$) is defined as:}
\begin{equation}\label{eq:mij_k}
    m_{ij} = \log_{10} \, \sum^K_{k=1} A_i^k \, q_{k,j}^Z ~.
\end{equation}
Each star $i$ has $K$ process amplitudes ($A^k_i$)
and each element $j$ has $K$ \add{metallicity dependent} process vector components ($q_{k,j}^{Z}$). 
\add{The $Z$ superscript denotes the dependence of the process vectors on metallicity, $Z$, taken to be [Mg/H].}
The $m_{ij}$ value is the model expected logarithmic abundance of element $j$ \add{relative to Hydrogen in star $i$, such that the observed abundance can be expressed}
\begin{equation}\label{eq:xh}
    [\text{X}_j/\text{H}]_i = m_{ij} + \text{noise},
\end{equation}
\add{where ``noise'' represents observational noise and/or other sources of intrinsic abundance scatter that are not included in this model.}

\add{For example, in the $K=2$ case for a star ($i=40$) with a O abundance of $[\text{O}/\text{H}] =0.155$ and $\mgh = 0.122$, \name{} fits $q_{\text{CC,O}}^{\mgh=0.12} = 0.91$ and $q_{\text{Ia,O}}^{\mgh=0.12} = 0.13$. This indicates that at $\mgh = 0.12$, stellar O is dominantly produced by the prompt (CCSN) process. For this star, \name{} also derives $A_{40}^{\rm CC} = 1.3$ and $A_{40}^{\rm Ia} = 1.2$ from the abundances of all 15 elements. Summing the products of the corresponding process vector component and process amplitude, we find $m_{40,\text{O}} = \log_{10} ( 1.36 ) = 0.135$. \name{} predicts the star's O abundance to 0.02 dex, the typical scale of observational noise on [O/H] in APOGEE. For more detailed examples from a similar model, see Section 2 of W22.}
\ejg{Hogg let me know what you think. We don't introduce the CC and Ia notation until later in this section so I think if it stays it needs more clarification. I lean towards just sending people to Section 2 and Figures 2 and 3 of W22. }

In \name{}, we put forth the following set of assumptions:

\paragraph{1. $K$ processes}
All elements on the periodic table are produced by a combination of nucleosynthetic processes such as CCSN, SNIa, AGB stars, and merging neutron stars \citep{johnsonja2020}. The majority of $\alpha$, light odd-Z, and Fe-peak elements (the elements observed by APOGEE) are dominantly produced by $K=2$ sources, with one being a prompt process or mix of prompt processes, and one being a delayed process or a mix of delayed processes. This is substantiated by theoretical yields \citep[e.g.,][]{anderson2019, rybizki2017} and past successful data-driven models (e.g., \citealp{ness2019}, G22, W22, \citealp{ting2022, ratcliffe2023}). In this paper we therefore assume that $K \geq 2$, though \name{} could in principle be implemented with $K=1$.

@@

In \name{}, this assumption is enforced by fixing the Mg process vector components such that
\begin{equation}\label{eq:qcc_mg_solar}
    q_{{\rm CC, Mg}}^{\,Z} = 1, \quad q_{k>1,{\rm Mg}}^{\,Z} = 0
\end{equation}
at all metallicities.

\paragraph{5. Fe production}
Fe is produced through a combination of a prompt and delayed process. Because the delayed process is likely dominated by SNIa \citep[e.g.,][]{andrews2017}, we label the delayed process with ``Ia''\footnote{While other enrichment channels with similar timescales may be included in the respective processes, the ``CC'' and ``Ia'' naming convention conforms to the choices in W22 and G22, and avoids the possible confusion of process numbers (1 and 2) with supernova type (II and Ia).}.
While the prompt process constraint (Mg) is grounded in nucleosynthesis theory, there is no equivalent nucleosynthesis fact to constrain the delayed process.
To break model degeneracies, we also fix the Fe process vector components such that

\begin{equation}\label{eq:qcc_fe_solar}
    q_{{\rm CC, Fe}}^{\,Z} = 0.4, \quad q_{{\rm Ia, Fe}}^{\,Z} = 1 - q_{{\rm CC, Fe}}^{\,Z} \quad q_{k>2,{\rm Fe}}^{\,Z} = 0
\end{equation}
at all metallicities. 
This assumption places a star with purely prompt enrichment on the low-metallicity $\mgfe$ plateau near 0.4 dex, in agreement with APOGEE observations but in contention with recent results from \citet{conroy2022} which place the $\mgfe$ plateau near 0.6 dex.
We explore the impact of different $\qccFe$ assumptions in Section~\ref{subsec:qccFe}.

@@

\paragraph{7. APOGEE abundances and uncertainties}
We assume that the APOGEE abundances and uncertainties can be used for this project.
This is not the same as assuming that they are correct, but rather that it is possible and useful to build an interpretable model to explain them. We describe the potential data systematics in Section~\ref{sec:data}.
\add{For our purposes, we care merely about the statistical observational errors rather than systematics that arise from imperfect modeling of the spectra such as NLTE effects, though differential systematics across the sample can artificially add abundance scatter. The actual derived values of $q_{k,j}^{Z}$ will be affected by systematic offsets in the abundances. We add a softening parameter $Q$ (Equation~\ref{eq:inflate_ivar}) to allow for the possibility that APOGEE observational errors are underestimated or that there is intrinsic scatter around the \name{} predictions.}
% The APOGEE errors may be slightly underestimated for some element measurements, so we add a softening parameter $Q$ to inflate the uncertainties on abundance-outlier stars (see Equations~\ref{eq:chi2} and ~\ref{eq:inflate_ivar}).

\paragraph{8. Robust likelihood function}
The observed value of [X/H] can be described as the $K$ process expected value plus observational noise and/or other sources of intrinsic abundance scatter, \add{as described by Equation~\ref{eq:xh}.}
The expression in this equation can be thought of as the key assumption underlying our likelihood function.
In detail the (negative two times the) log likelihood function is given by a chi-squared ($\chi^2$) objective
\begin{equation}\label{eq:chi2}
    \chi^2 = \sum_{ij}\frac{1}{\sigma_{ij}^2} \, (\xh_{ij} - m_{ij})^2 ~,
\end{equation}
where $1/\sigma_{ij}^2$ is the (robust; see below) inverse variance on measurement $ij$.

Because we don't want to be too drawn or influenced by outlier points, we don't use the observed errors $\sigma_{{\rm obs},ij}$ in the likelihood, but instead we soften them in the spirit of iteratively reweighted least squares (\citealt{irls}):
\begin{equation}\label{eq:inflate_ivar}
    \frac{1}{\sigma_{ij}^2} = \frac{Q^2/\sigma_{{\rm obs},ij}^2}{Q^2 + (\xh_{ij} - m_{ij})^2 /\sigma_{{\rm obs},ij}^2 },    
\end{equation}
where $Q$ is a softening parameter. \add{Our results are largely insensitive to the choice of $Q$ (e.g., $1 < Q < 10$), so we choose to set $Q=5$. Very small values (e.g., $Q=.1$) will erase some of the abundance structure and produce poorer fits.}

\paragraph{9. Implementation and optimization} 
With the above assumptions in place, the likelihood function can be optimized to a set of stellar abundances.
The model is initialized at the Mg and Fe process vector components from Equations~\ref{eq:qcc_mg_solar} --~\ref{eq:qcc_fe_z}. It subsequently optimizes the process amplitudes (dubbed the $A$-step) using only Mg and Fe at fixed process vector components, and then optimizes the process vector components (dubbed the $q$-step) for all elements at fixed process amplitudes.
The $A$-step and $q$-step are alternated, repeating 48 rounds of optimization, \add{in the $K=2$ case,} and updating the best-fit parameters when the objective function improves. We find few differences in the best-fit parameters when we decrease the number of iterations to 32, indicating that the model quickly finds a good solution.  
In detail the optimizations are performed with a nonlinear $\chi^2$ minimization algorithm (Gauss--Newton nonlinear least-squares) from \texttt{jaxopt}\footnote{\url{https://jaxopt.github.io/}}.

\bigskip
\name{} mirrors the two-proccess model from prior work (G22, W22) but, unless otherwise noted, the assumptions are weaker, there is a likelihood function in play, and the implementation is more general.
In particular, we don't assume anything about the relationships between the process vector components and the morphologies of observed element-abundance ratio diagrams.
